Mergeable Types

### 2017-02-28 paper/examples.tex

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} described in the introduction.  An alternative
instantiation of \C{MList} is for a quantity type
(\C{Qty}) that supports \C{add} and \C{subtract} operations, and
defines its \C{merge} operation a little differently:
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.45in}
\end{wrapfigure}
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is a shopping list application that allows its users
to collaboratively build a shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
merge function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.35in}
\end{wrapfigure}
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is certainly
allowed).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.
The examples described above make certain assumptions about the
concurrency model and the underlying system, such as the existence of
a single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

>>>>

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and \C{Int}
types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same as
\C{Counter.merge} described in the introduction.  An alternative instantiation
of \C{MList} is for a quantity type (\C{Qty}) that supports \C{add} and
\C{subtract} operations, and defines its \C{merge} operation a little
differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.45in}
\end{wrapfigure}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has the same
semantics under a sequential execution. Their differences however manifest
under a concurrent execution, where both adopt different methods to reconcile
concurrent updates.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is a shopping list application that allows its users
to collaboratively build a shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
merge function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is certainly
allowed).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
\begin{wrapfigure}{r}{.35\textwidth}
  \centering
  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.35in}
\end{wrapfigure}
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.
The examples described above make certain assumptions about the
concurrency model and the underlying system, such as the existence of
a single lowest common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

### 2017-02-28 paper/implementation.tex

Concurrent operations are tracked within the notion of branches, but
importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are defined
over objects on Irmin's content-addressable heap, \name's merge functions are
defined over OCaml types. We address this representational mis-match with the
help of OCaml's PPX metaprogramming support to derive bi-directional
transformations between objects on OCaml and Irmin heaps. We also derive the
various serialization functions required Irmin. This address the representation
mismatch between the application and storage layers. Synchronization between
replicas is performed using Git's notion of pushing and pulling updates from
remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred
between replicas), compression and end-to-end encrypted communication between
the replicas.

>>>>

Concurrent operations in Irmin are tracked within the notion of branches,
allowing the programmer to explicitly merge the branches on demand. \name
realizes the concurrent programming model on top of the notion of branches;
each replica operates on its own branch, and thus is isolated from actions on
other replicas. \name monad conceals the branching structure, but also
transparently performs the necessary merges to obtain a history graph with a
unique lowest common ancestor.

Importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are defined
over objects on Irmin's content-addressable heap, \name's merge functions are
defined over OCaml types. We address this representational mis-match with the
help of OCaml's PPX metaprogramming support to derive bi-directional
transformations between objects on OCaml and Irmin heaps. We also derive the
various serialization functions required Irmin. This address the representation
mismatch between the application and storage layers. Synchronization between
replicas is performed using Git's notion of pushing and pulling updates from
remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred
between replicas), compression and end-to-end encrypted communication between
the replicas.

### 2017-02-28 paper/intro-sj.tex

On the other hand, a functional programmer is likely to approach the
problems very differently.  Rather than thinking about a low-level
operational variant of distributed state given in terms of explicit
effects as described above, a declarative interpretation of a
decentralized computation could be naturally described in terms of a
version tree whose root contains the initial counter value, and whose
branches represent different versions maintained by different nodes,
with the state produced by a counter's operations on a node recorded
along the node's local branch.  Now, a globally consistent view of a
counter can be explained simply in terms of \emph{merge} operations of
different branch tips.  Treating replication as merging yields a
counter implementation that bears strong similarity to the original
sequential one:
  \begin{ocaml}
    module Counter = struct
      include Counter
      type mergeable_t = t [@@deriving persistence]
      let merge lca v1 v2 = lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
data type is thus given in terms of multi-versions of a program state
impicitly associated with the different nodes that comprise a
distributed application and a logical notion of a \emph{master}
representing some consistent, but not necessarily current, view of
global state.  
\begin{figure}
  \begin{center}
    \includegraphics[scale=0.4]{Figures/dali-counter}
  \end{center}
  \caption{\small A multi-version counter can be expressed as a
    non-planar graph, consisting of a logical master that defines a
    notion of a globally consistent state, and a collection of
    versions, each associated with a different node in a distributed
    system.  Versions periodically merge their state with the master.
    In the figure, when the local state of node 2, labeled 1, merges
    with the master whose counter state is 15, a new counter value is
    produced.  This value takes into account the global state (10)
    from which the versions under merge were initiated to yield a new
    global state of 6.  Observe that there is no synchronization or
    coordination among versions.  In the figure, red circles represent
    state produced through coordination with the master, while blue
    circles represent local counter state produced by applying counter
    operations on a node-local version.  }
\end{figure}

>>>>

On the other hand, a functional programmer is likely to approach the
problems very differently.  Rather than thinking about a low-level
operational variant of distributed state given in terms of explicit
effects as described above, a declarative interpretation of a
decentralized computation could be naturally described in terms of a
version tree whose root contains the initial counter value, and whose
branches represent different versions maintained by different nodes,
with the state produced by a counter's operations on a node recorded
along the node's local branch.  Now, a globally consistent view of a
counter can be explained simply in terms of \emph{merge} operations of
different branch tips.  Treating replication as merging yields a
counter implementation that bears strong similarity to the original
sequential one:
  \begin{ocaml}
    module Counter = struct
      include Counter
      type mergeable_t = t [@@deriving persistence]
      let merge lca v1 v2 = lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (lowest common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
data type is thus given in terms of multi-versions of a program state
impicitly associated with the different nodes that comprise a
distributed application and a logical notion of a \emph{master}
representing some consistent, but not necessarily current, view of
global state.
\begin{figure}
  \begin{center}
    \includegraphics[scale=0.4]{Figures/dali-counter}
  \end{center}
  \caption{\small A multi-version counter can be expressed as a
    non-planar graph, consisting of a logical master that defines a
    notion of a globally consistent state, and a collection of
    versions, each associated with a different node in a distributed
    system.  Versions periodically merge their state with the master.
    In the figure, when the local state of node 2, labeled 1, merges
    with the master whose counter state is 15, a new counter value is
    produced.  This value takes into account the global state (10)
    from which the versions under merge were initiated to yield a new
    global state of 6.  Observe that there is no synchronization or
    coordination among versions.  In the figure, red circles represent
    state produced through coordination with the master, while blue
    circles represent local counter state produced by applying counter
    operations on a node-local version.  }
\end{figure}

### 2017-02-28 paper/intro.tex

Framing replication as merging leads to a counter implementation that
bears strong similarity to the original sequential one:
  \begin{ocaml}
    module Replicated_Counter = struct
      include Counter
      let merge lca v1 v2 =
         lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (least common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
implicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

>>>>

Framing replication as merging leads to a counter implementation that
bears strong similarity to the original sequential one:
  \begin{ocaml}
    module Replicated_Counter = struct
      include Counter
      let merge lca v1 v2 =
         lca + (v1 - lca) + (v2 - lca)
    end
  \end{ocaml}
The role of \C{lca} (lowest common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor.  This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
implicitly associated with the different nodes that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

### 2017-02-28 paper/mergeable-types.tex
fix wrapfig

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the computation's branching structure. The merge
function, as described above, however assumes that concurrent versions
are semantic successors of their LCA.  It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing an ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts}
\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{wrapfigure}
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, the merge function has to enforce the relationship
explicitly. In other words, the result of merging  a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
$D:t$. This intuition is expressed visually in
Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
guarantee that a sequence of values read by a thread from its branch
is sensible, as per the data type semantics:

@@

\noindent Rather than specifying merge ($M$) in terms of an LCA as we
did in describing our programming model,
Def.~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of
certain nonsensical merge functions, and thus enforce a useful
branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}  
Such a merge function allows the client to witness a violation of
branch-local consistency. For example, the client may
\C{sync} the counter, whose current local value is $10$, and obtain
$0$ as a result, thus witnessing a violation of a counter's monotonicity
property. Fortunately, Def.~\ref{def:mergeable-type} disallows this
semantics. It is impossible to define the function $M$ for the counter
data type that violates the monotonicity invariant, because no counter
morphism violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms
returned by $M$ map respective concurrent versions to the same
value. The \rulelabel{E-Pull-Wait} rule of the operational semantics
(Fig.~\ref{fig:opsem}) can be extended to this effect:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  t\neq t' \spc
% \under{H}{v' \mbleto v} \spc
% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc 
  v' \not\preceq v \spc
  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
  Q'(v) = P'(v')
}
{
  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
}
\end{array}
\end{smathpar}

@@

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which (implicitly) computes the
edit distance between an object $A:t$ and its semantic successor
$B:t$.  The sequence of primitive morphisms it returns ($P^*$) is
called an edit sequence. Applying the edit sequence on $A$ produces
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
$B$ (i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an operational
transformation of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$.

>>>>

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their lowest common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the computation's branching structure. The merge
function, as described above, however assumes that concurrent versions
are semantic successors of their LCA.  It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing an ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
\begin{wrapfigure}{L}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts}
\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{wrapfigure}
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, the merge function has to enforce the relationship
explicitly. In other words, the result of merging  a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
$D:t$. This intuition is expressed visually in
Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
guarantee that a sequence of values read by a thread from its branch
is sensible, as per the data type semantics:

@@

\noindent Rather than specifying merge ($M$) in terms of an LCA as we
did in describing our programming model,
Def.~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of
certain nonsensical merge functions, and thus enforce a useful
branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to witness a violation of
branch-local consistency. For example, the client may
\C{sync} the counter, whose current local value is $10$, and obtain
$0$ as a result, thus witnessing a violation of a counter's monotonicity
property. Fortunately, Def.~\ref{def:mergeable-type} disallows this
semantics. It is impossible to define the function $M$ for the counter
data type that violates the monotonicity invariant, because no counter
morphism violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms
returned by $M$ map respective concurrent versions to the same
value. The \rulelabel{E-Pull-Wait} rule of the operational semantics
(Fig.~\ref{fig:opsem}) can be extended to this effect:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  t\neq t' \spc
  \under{H}{v' \mbleto v} \spc
% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc
  v' \not\preceq v \spc
  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
  Q'(v) = P'(v')
}
{
  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
}
\end{array}
\end{smathpar}

@@

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which (implicitly) computes the
edit distance between an object $A:t$ and its semantic successor
$B:t$.  The sequence of primitive morphisms it returns ($P^*$) is
called an edit sequence. Applying the edit sequence on $A$ produces
\begin{wrapfigure}{L}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
$B$ (i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an operational
transformation of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$.
